An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-07 main.tex
Tweak intro, clean up bibliography

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  A common problem in string constraint solvers is computing the Parikh image, a
   set of linear equations that describe all possible combinations of character
   counts in strings of a given language. Automata-based string solvers
   frequently need to compute the Parikh image of large products (intersections)
   of automata, which in many operations is both prohibitively slow and
   memory-intensive. We contribute a novel understanding of how Parikh maps can
   be tackled as a constraint-solving problem to solve real-world constraints
   stemming from functions on regular languages, most notably the length
   constraint. Furthermore, we show how this formulation can be efficiently
   implemented as a calculus, \Calculus{}, in an automated theorem prover
   supporting Presburger logic. We also provide an efficient method for deriving
   the Presburger formula of a given Parikh image, \Fudge{which is good for
   something} based on techniques for quantifier elimination. We implement
   \Calculus{} in a tool called \Catra{}, and evaluate it on constraints
   produced by the \OstrichPlus{} string constraint solver when solving
   \Fudge{standard string constraint benchmarks involving string lengths}. We
   show that our solution strictly outperforms the standard eager approach
   described in~\citeauthor{generate-parikh-image} as well as the
   over-approximating method recently described
   by~\citeauthor{approximate-parikh}, and for realistic timeouts for constraint
   solving also the~\Nuxmv{} model checker.
\end{abstract}

@@

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, the method can be used iteratively to tackle smaller chunks
of the product incrementally, thereby avoiding running out of memory at all.

@@

In addition to the decision problem of determining whether an intersection of
Parikh automata is empty and the satisfaction problem of finding a value from
the intersection if it is not, \Catra{} is able to efficiently generate a
quantifier-free Presburger formula representing an intersection of Parikh
automata, something beyond the reach of general model checkers like~\Nuxmv.
\Fudge{This is useful because the moon is made of cheese.}

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in a tool that we call \Catra{}. \Catra also supports using the
approximate method of \cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra{}, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata satisfaction problems generated
\Fudge{by \OstrichPlus{}}.

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute (a homomorphism on) the
Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image. \Fudge{Why is this good?}
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future study.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\end{itemize}

@@

\begin{table}
\begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
  Name & Rule & Prerequisites\\
  \midrule

  % EXPAND
  \Expand & 
    \inferrule
  {\Connected(\Automaton, \Filter) \land \FlowEq(\Automaton, \Filter) \land \MonoidElement = \sum_{\Transition \in \Transitions_\Automaton} \Filter(\Transition) \cdot \Map(\Transition), \SomeInequalities, \SomeClause}
  {\SinglePredicateInstance, \SomeInequalities, \SomeClause} & 
  None \\[3ex]

  % SPLIT
  \Split & 
  \inferrule{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) = 0 \mid \Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) > 0}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  if $\Transition \in \Transitions_\Automaton$ \\[3ex]

  % PROPAGATE
  \Propagate &
  \inferrule{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause, \Filter(\Transition) = 0}{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause} &
  if $t = \FromLabelTo{\State}{}{\State'} \in \Transitions_\Automaton, \SeparatingCut(C, \Automaton, \State)$ \\[3ex]

@@

  \begin{table}[h]
    \begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
      Name & Rule & Prerequisites\\
      \midrule

      % EXPAND
      \ExpandM & 
      \inferrule
      {
        {\begin{matrix}
          \Set{ 
            \MonoidElement = \sum_{\Transition \in \Transitions_{\Automaton_i}} \Filter(\Transition) \cdot \Map(\Transition)
          \SuchThat \Automaton_i \in \Automaton_1,\ldots,\Automaton_k}, \\
          \Set{\FlowEq(\Automaton_i), \Connected(\Automaton_i) \SuchThat \Automaton_i \in \Automaton_1,\ldots,\Automaton_k}, \\
          \ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \\
          \SomeInequalities, \SomeClause
        \end{matrix}}
        }
      {\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} & 
      None \\
      \Materialise &
      \inferrule
      {    \BindingSum(\Automaton', \Filter),\ImagePredicate{\Automaton'\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
      {\ImagePredicate{\Automaton_1\times\Automaton_2\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} &
      $\Automaton' = \Automaton_1\times\Automaton_2$ \\
      \bottomrule
      \end{tabular}
      \caption{Additional derivation rules for products of arbitrarily many automata.}\label{tbl:rules:multi}
    \end{table}

@@

\begin{figure}[t]
  \begin{minipage}[b]{0.49\linewidth}
    \centering
      \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
      \caption{The division of statuses per backend.}
      \label{fig:solve-division}
      \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly more in total than lazy.}
  \end{minipage}
  \begin{minipage}[b]{0.49\linewidth}
    \centering
      \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
      \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
      \label{fig:runtime-boxplot}
      \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{minipage}
  \vspace{0.5cm}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv on shorter runtimes, but is overtaken by just before the 60-second timeout used in these experiments.}
  \end{minipage}
\end{figure}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  A common problem in string constraint solvers is computing the Parikh image, a
   set of linear equations that describe all possible combinations of character
   counts in strings of a given language. Automata-based string solvers
   frequently need to compute the Parikh image of large products (intersections)
   of nondeterministic automata, which in many operations is both prohibitively
   slow and memory-intensive. We contribute a novel understanding of how Parikh
   maps can be tackled as a constraint solving problem to solve real-world
   constraints stemming from functions on regular languages, most notably the
   length constraint. Furthermore, we show how this formulation can be
   efficiently implemented as a calculus, \Calculus{}, in an automated theorem
   prover supporting Presburger logic.  We implement \Calculus{} in a tool
   called \Catra{}, and evaluate it on constraints produced by the
   \OstrichPlus{} string constraint solver when solving \Fudge{standard string
   constraint benchmarks involving string lengths}. We show that our solution
   strictly outperforms the standard eager approach described
   in~\citeauthor{generate-parikh-image} as well as the over-approximating
   method recently described by~\citeauthor{approximate-parikh} by a wuide
   marigin, and for realistic timeouts for constraint solving also the~\Nuxmv{}
   model checker.
\end{abstract}

@@

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, the method can be used iteratively to tackle smaller chunks
of the product incrementally, thereby decreasing the memory footprint.

@@

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in the tool \Catra{}. \Catra also supports using the
approximate method of \cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra{}, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata satisfaction problems generated
\Fudge{by \OstrichPlus{}}.

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute (a homomorphism on) the
Parikh image of products of Parikh automata.
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future study.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\end{itemize}

@@

\begin{table}
\begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
  Name & \text{Rule} & Prerequisites\\
  \midrule

  % EXPAND
  \Expand & 
    \inferrule
  {\Connected(\Automaton, \Filter) \land \FlowEq(\Automaton, \Filter) \land \MonoidElement = \sum_{\Transition \in \Transitions_\Automaton} \Filter(\Transition) \cdot \Map(\Transition), \SomeInequalities, \SomeClause}
  {\SinglePredicateInstance, \SomeInequalities, \SomeClause} & 
  None \\[4ex]

  % SPLIT
  \Split & 
  \inferrule{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) = 0 \mid \Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) > 0}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  if $\Transition \in \Transitions_\Automaton$ \\[4ex]

  % PROPAGATE
  \Propagate &
  \inferrule{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause, \Filter(\Transition) = 0}{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause} &
  if $t = \FromLabelTo{\State}{}{\State'} \in \Transitions_\Automaton, \SeparatingCut(C, \Automaton, \State)$ \\[4ex]

@@

  \begin{table}[t]
    \begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
      Name & \text{Rule} & Prerequisites\\
      \midrule

      % EXPAND
      \ExpandM & 
      \inferrule
      {
        {\begin{matrix}
          \Set{ 
            \MonoidElement = \sum_{\Transition \in \Transitions_{\Automaton_i}} \Filter(\Transition) \cdot \Map(\Transition)
          \SuchThat \Automaton_i \in \Automaton_1,\ldots,\Automaton_k}, \\
          \Set{\FlowEq(\Automaton_i), \Connected(\Automaton_i) \SuchThat \Automaton_i \in \Automaton_1,\ldots,\Automaton_k}, \\
          \ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \\
          \SomeInequalities, \SomeClause
        \end{matrix}}
        }
      {\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} & 
      None \\[5ex]
      \Materialise &
      \inferrule
      {    \BindingSum(\Automaton', \Filter),\ImagePredicate{\Automaton'\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
      {\ImagePredicate{\Automaton_1\times\Automaton_2\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} &
      $\Automaton' = \Automaton_1\times\Automaton_2$ \\
      \bottomrule
      \end{tabular}
      \caption{Additional derivation rules for products of arbitrarily many automata.}\label{tbl:rules:multi}
    \end{table}

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly more in total than lazy.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{0.75\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv on shorter runtimes, but is overtaken by just before the 60-second timeout used in these experiments.}
  \end{subfigure}
\end{figure}
